---
title: "Assignment 07"
author: "Johnny Willing"
format: html
embed-resources: true
execute: 
  warning: false
  message: false
---

## Set up

```{r}
library(tidyverse)
library(tidymodels)
library(corrr)
library(doParallel)
library(future)

setwd("/Users/JohnnyWilling/Documents/Grad school/Spring 2026/PPOL 6891 Advanced Data Science/Assignments/ppol6819.assignment07")

nreturn_train <- read_csv("data/assignment_modeling.csv")

```

## Exploratory Data Analysis

```{r}
glimpse(nreturn_train)
colSums(is.na(nreturn_train))

# Looks like there are a lot of variables that could cause data leakage, will need to drop. And the pct_TEA vars have a lot of missingness
data_leak_vars <- c("Low_Response_Score", "Self_Response_Rate_ACS_13_17", "pct_Deletes_CEN_2010", "pct_Census_UAA_CEN_2010", "pct_Mailback_Count_CEN_2010",
                    "pct_TEA_MailOutMailBack_CEN_2010", "pct_TEA_Update_Leave_CEN_2010", "pct_Vacants_CEN_2010")

# There are also a lot of duplicated variables - reporting the exact same thing, just as percentage vs count or from decennial census vs ACS
# Keeping percents and drop any variable for which there exists another variable that has the same name, plus the prefix "pct_"
# Keeping ACS data since it's more recent, and dropping decennial census duplicates
varnames <- names(nreturn_train)
pct_vars <- varnames[str_detect(varnames, "^pct_")]
nonpct_vars <- str_remove(pct_vars, "^pct_")
acs_vars <- varnames[str_detect(varnames, "ACS_13_17$")]
cen_dup_vars <- str_c(str_remove(acs_vars, "ACS_13_17$"), "CEN_2010")
dup_vars <- varnames[varnames %in% c(nonpct_vars, cen_dup_vars)]

vars_to_drop <- c(data_leak_vars, dup_vars)

# Get a sense of distribution of non-return rate
nreturn_train |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram()

# Check correlations between non-return rate and other vars
nreturn_train |>
  select(-all_of(vars_to_drop)) |>
  select(where(is.numeric)) |>
  correlate() |>
  focus(non_return_rate) |>
  arrange(desc(abs(non_return_rate))) |>
  head(30) |>
  ggplot(aes(x = term, y = non_return_rate)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = 75)) +
  labs(x = "Variable", y = "Correlation with non-return rate")

# See if rates vary significantly across states
nreturn_train |>
  group_by(State_name) |>
  summarize(non_return_rate = mean(non_return_rate)) |>
  ggplot(aes(x = reorder(State_name, non_return_rate), y = non_return_rate)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = 75)) +
  labs(x = "State", y = "Average non-return rate")

# Some dollar variables are coded as characters, converting to numeric. Outside of recipe to save time during tuning/training
nreturn_train <- nreturn_train |>
  mutate(across(c(Aggregate_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17,
      avg_Agg_HH_INC_ACS_13_17, avg_Agg_House_Value_ACS_13_17), 
      ~ as.numeric(gsub("[\\$,]", "", .))))


```

## Set up folds
```{r}
set.seed(20260420)
folds <- vfold_cv(data = nreturn_train, v = 3)

```

## Tuning models

### Lasso Regression
```{r}
# parallelize
plan(multisession, workers = parallel::detectCores() - 2)

# recipe
lasso_rec <- recipe(formula = non_return_rate ~ ., data = nreturn_train) |>
  step_rm(all_of(vars_to_drop)) |> # drop the duplicate/data leak variables
  step_zv(all_predictors()) |> # drop vars with zero variance
  add_role(GIDTR, County_name, new_role = "id") |> 
  step_rm(has_role("id")) |> # drop ID vars
  step_impute_median(all_numeric_predictors()) |> # impute median for missing values
  step_log(Med_HHD_Inc_ACS_13_17, Aggregate_HH_INC_ACS_13_17, 
           Med_House_Value_ACS_13_17, Aggr_House_Value_ACS_13_17, 
           avg_Agg_HH_INC_ACS_13_17, avg_Agg_House_Value_ACS_13_17) |> # logging dollar value vars such as income, as they're likely very skewed
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

# model
lasso_mod <- linear_reg(penalty = tune(), mixture = 1) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "glmnet")

# tune penalty
lasso_grid <- grid_regular(penalty(), levels = 20)

# workflow
lasso_wf <- workflow() |>
  add_recipe(lasso_rec) |>
  add_model(lasso_mod)

# results
lasso_res <- lasso_wf |>
  tune_grid(resamples = folds,
            grid = lasso_grid,
            metrics = metric_set(yardstick::rmse, yardstick::mae))

lasso_res |>
  collect_metrics()

lasso_res |>
  autoplot()

lasso_res |>
  show_best()

```

### Regression Tree
```{r}
# recipe
tree_rec <- recipe(formula = non_return_rate ~ ., data = nreturn_train) |>
  step_zv(all_predictors()) |> # drop vars with zero variance
  add_role(GIDTR, County_name, new_role = "id") |> 
  step_rm(has_role("id")) |> # drop ID vars
  step_impute_median(all_numeric_predictors()) |> # impute median for missing values
  step_log(Med_HHD_Inc_ACS_13_17, Aggregate_HH_INC_ACS_13_17, 
           Med_House_Value_ACS_13_17, Aggr_House_Value_ACS_13_17, 
           avg_Agg_HH_INC_ACS_13_17, avg_Agg_House_Value_ACS_13_17) |> # logging dollar value vars such as income, as they're likely very skewed
  step_rm(all_of(vars_to_drop)) |> # drop the duplicate/data leak variables
  step_dummy(all_nominal_predictors())

# model
tree_mod <- decision_tree(cost_complexity = tune(),
    tree_depth = tune(), min_n = tune()) |>
  set_engine(engine = "rpart") |>
  set_mode(mode = "regression")

# tune hyperparameters - cost complexity, tree depth, and minimum node size
tree_grid <- grid_max_entropy(
  cost_complexity(range = c(-5, -1)),
  tree_depth(),
  min_n(),
  size = 40
)

# workflow
tree_wf <- workflow() |>
  add_recipe(tree_rec) |>
  add_model(tree_mod)

# results
tree_res <- tree_wf |>
  tune_grid(resamples = folds,
            grid = tree_grid,
            metrics = metric_set(yardstick::rmse, yardstick::mae))

tree_res |>
  collect_metrics()

tree_res |>
  show_best()

tree_res |>
  autoplot()

```

### Random Forest
```{r}
# recipe
rf_rec <- recipe(formula = non_return_rate ~ ., data = nreturn_train) |>
  step_zv(all_predictors()) |> # drop vars with zero variance
  add_role(GIDTR, County_name, new_role = "id") |> 
  step_rm(has_role("id")) |> # drop ID vars
  step_impute_median(all_numeric_predictors()) |> # impute median for missing values
  step_log(Med_HHD_Inc_ACS_13_17, Aggregate_HH_INC_ACS_13_17, 
           Med_House_Value_ACS_13_17, Aggr_House_Value_ACS_13_17, 
           avg_Agg_HH_INC_ACS_13_17, avg_Agg_House_Value_ACS_13_17) |> # logging dollar value vars such as income, as they're likely very skewed
  step_rm(all_of(vars_to_drop)) |> # drop the duplicate/data leak variables
  step_corr(all_numeric_predictors(), threshold = 0.9) |> # get rid of highly correlated predictors to save computing time
  step_dummy(all_nominal_predictors())

# model
rf_mod <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 100
  ) |>
  set_mode(mode = "regression") |>
  set_engine(
    engine = "ranger", 
    importance = "impurity",
    num.threads = 4
  )

# tune hyperparameters - mtry and minimum node size
rf_grid <- grid_regular(
  mtry(range = c(5, 35)),
  min_n(range = c(5, 30)),
  levels = 5
)

# workflow
rf_wf <- workflow() |>
  add_recipe(rf_rec) |>
  add_model(rf_mod)


# results
set.seed(20260420)
rf_res <- rf_wf |>
  tune_grid(resamples = folds,
            grid = rf_grid,
            metrics = metric_set(yardstick::rmse, yardstick::mae))

rf_res |>
  collect_metrics()

rf_res |>
  show_best()

rf_res |>
  autoplot()

```

Random forest had better performance than Lasso and a single regression tree; I'll now use this model for the final fit on the entire training dataset. For tuning, I used 100 trees to save time, but I will use 500 for training the final model.

## Training final model on full training set
```{r}
# select best parameters from random forest results
best_rf <- rf_res |>
  select_best(metric = "rmse")

# create best model from random forest results, updating trees from 100 to 500
final_mod <- rand_forest(
  mtry = best_rf$mtry,
  min_n = best_rf$min_n,
  trees = 500
  ) |>
  set_mode(mode = "regression") |>
  set_engine(
    engine = "ranger", 
    importance = "impurity",
    num.threads = 4
  )

# final workflow
final_wf <- workflow() |>
  add_recipe(rf_rec) |>
  add_model(final_mod)

# fit the final model on entire training set
set.seed(20260420)
final_fit <- final_wf |>
  fit(data = nreturn_train)

```

# Generate final predictions
```{r}
nreturn_test <- read_csv("data/assignment_implementation.csv") |>
  mutate(across(c(Aggregate_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17,
      avg_Agg_HH_INC_ACS_13_17, avg_Agg_House_Value_ACS_13_17), 
      ~ as.numeric(gsub("[\\$,]", "", .)))) # fixing dollar variables are coded as characters

predictions <- bind_cols(
  nreturn_test,
  predict(final_fit, new_data = nreturn_test)
) |>
  select(GIDTR, .pred)

```
